Exploring the internal language of toposes

### 2018-06-28 slides-como2018.tex
More slides

\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\renewcommand{\AA}{\mathbb{A}}
\renewcommand{\C}{\mathcal{C}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\renewcommand{\G}{\mathcal{G}}
\newcommand{\GG}{\mathbb{G}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\ZZ}{\mathbb{Z}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\fff}{\mathfrak{f}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\Zar}{\mathrm{Zar}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\newcommand{\Sch}{\mathrm{Sch}}
\newcommand{\Aff}{\mathrm{Aff}}
\newcommand{\LRS}{\mathrm{LRS}}
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\Spec}{\mathrm{Spec}}
\newcommand{\lra}{\longrightarrow}
\newcommand{\RelSpec}{\operatorname{Spec}}
\renewcommand{\_}{\mathpunct{.}}
\newcommand{\?}{\,{:}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\ull}[1]{\underline{#1}}
\newcommand{\affl}{\ensuremath{{\ull{\AA}^1}}}
\newcommand{\Ll}{\text{iff}}
\newcommand{\inv}{inv.\@}
\newcommand{\seq}{\vdash_{\!\!\!\vec x}}

@@

@@

\begin{frame}[c]
  \centering
  \medskip
  \hspace*{-3.9em}%
%  \includegraphics[height=0.3\textheight]{vichy-2}%
%  \includegraphics[height=0.3\textheight]{vichy-1}%
%  \includegraphics[height=0.3\textheight]{vichy-4}%
%  \includegraphics[height=0.3\textheight]{vichy-5}%

  \hil{Exploring the internal language of toposes} \\

@@

  \bigskip

@@

      \hil{Alternate universes}
      \medskip

    \end{column}
    \hspace*{0.5em}
    \begin{column}{0.25\textwidth}
      \centering
      \bignumber{2}

      \hil{Applications}
      \medskip

@@

      \hil{Vision for the future}
      \medskip

@@

  \scriptsize
  Ingo Blechschmidt (MPI Leipzig) \\
  Toposes in Como \\
  June 29th, 20s18
  \par
\end{frame}

@@

  % XXX: \includegraphics[height=3em]{generic-freeness}
  \vspace*{-1em}

@@

  \pause\raggedright
  \textbf{Proof.} See [Stacks Project, Tag 051Q].
\end{frame}

@@

  \vspace*{-1em}
  \begin{columns}
    \def\insertblocktitle{}
    \begin{column}{0.3\textwidth}\usebeamertemplate{block begin}
      \centering
      $\Set \models \varphi$ \\
      ``$\varphi$ holds in the \\ usual sense.''
      \usebeamertemplate{block end}\end{column}

    \begin{column}{0.3\textwidth}\usebeamertemplate{block begin}
      \centering
      $\Sh(X) \models \varphi$ \\
      ``$\varphi$ holds continuously.''
      \usebeamertemplate{block end}\end{column}

    \begin{column}{0.3\textwidth}\usebeamertemplate{block begin}
      \centering
      $\Eff \models \varphi$ \\
      ``$\varphi$ holds \\ computably.''
      \usebeamertemplate{block end}\end{column}
  \end{columns}
  \medskip

@@

\begin{frame}{Internalizing parameter-dependence}
  \justifying
  Let~$X$ be a space. A continuous family~$(f_x)_{x \in X}$ of continuous
  functions (that is, a continuous function~$f : X \times \RR \to \RR$;
  $f_x(a) = f(x,a)$) induces an endomorphism of the sheaf~$\C$ of continuous
  functions:
  \[
    \bar f : \C \longrightarrow \C,\
    \text{on $U$:}\ s \longmapsto (x \mapsto f_x(s(x))).
  \]
  \begin{itemize}
    \item $\Sh(X) \models \speak{The set $\C$ is the set of (Dedekind) reals}$.
    \item $\Sh(X) \models \speak{The function $\bar f : \RR \to \RR$ is continuous}$.
    \item Iff $f_x(-1) < 0$ for all $x$, then $\Sh(X) \models \bar f(-1) < 0$.
    \item Iff $f_x(+1) > 0$ for all $x$, then $\Sh(X) \models \bar f(+1) > 0$.
    \item Iff all $f_x$ are increasing, then $\Sh(X) \models \speak{$\bar f$ is increasing}$.
    \item \ \\[-1.2em]\mbox{Iff there is an open cover~$X = \bigcup_i U_i$ such
    that for each~$i$ there is a} \mbox{continuous function~$s : U_i \to \RR$
    with~$f_x(s(x)) = 0$ for all~$x \in U_i$,} then $\Sh(X) \models \exists s
    \? \RR\_ \bar f(s) = 0$.
  \end{itemize}
\end{frame}

\begin{frame}{The little Zariski topos}
  Let~$A$ be a ring. Its \hil{little Zariski topos} is equivalently
  \begin{enumerate}
    \item the classifying topos of \hil{local localizations} of $A$,
    \item the classifying topos of \hil{prime filters} of $A$,
    \item the locale given by the frame of \hil{radical ideals} of $A$,
    \item the topos of sheaves over the poset $A$ with $f \preceq g$ iff $f \in \sqrt{(g)}$ and with $(f_i \to f)_i$ deemed covering iff $f \in \sqrt{(f_i)_i}$ or
    \item the topos of sheaves over $\Spec(A)$.
  \end{enumerate}
  Its associated topological space of points is the \hil{classical spectrum}
  \[ \{ \fff \subseteq A \,|\, \text{$\fff$ prime filter} \} + \text{Zariski topology}. \]
  It has \hil{enough points} if the Boolean Prime Ideal Theorem holds.

@@

  \only<1>{\begin{center}
    \includegraphics[width=0.9\textwidth]{hartshorne-tilde-construction} \\
    \footnotesize
    Robin Hartshorne. Algebraic Geometry. 1977.
  \end{center}}

  \begin{columns}
    \begin{column}{0.5\textwidth}
      \begin{varblock}{\textwidth}{}
        \justifying
        Assuming the Boolean prime ideal theorem, a geometric sequent
        ``$\forall \ldots \forall\_ (\cdots \Longrightarrow \cdots\!\,)$'',
        where the two subformulas may not contain~``$\Rightarrow$'' and~``$\forall$'',
        holds for~$M^\sim$ iff it holds for all stalks~$M_\ppp$.
      \end{varblock}
    \end{column}

    \begin{column}{0.5\textwidth}
      \begin{varblock}{\textwidth}{}
        $M^\sim$ inherits any property of~$M$ which is \hil{localization-stable}.
      \end{varblock}

      \bigskip
      If $A$ is reduced ($x^n = 0 \Rightarrow x = 0$):

      \vspace*{-1em}
      \setbeamercolor{block body}{bg=red!30}
      \setbeamercolor{structure}{fg=purple}
      \begin{varblock}{\textwidth}{}
        $A^\sim$ is a \hil{field}
        (nonunits are zero).

@@

  \visible<3>{\begin{tikzpicture}[overlay]
    \draw[fill=white, draw=white, opacity=0.95] (-1,0) rectangle (\paperwidth,7.4);
    \node[anchor=south west,inner sep=0] (image) at (0,0.8) {\vbox{
      \centering
      \includegraphics[width=0.9\textwidth]{tierney-on-the-spectrum-of-a-ringed-topos} \\
      \footnotesize
      Miles Tierney. On the spectrum of a ringed topos. 1976.
    }};
  \end{tikzpicture}}
\end{frame}

@@

  \only<1>{\begin{columns}[t]
    \begin{column}[t]{0.50\textwidth}
      \centering

@@

      \justifying
      \textbf{Proof.} The matrix is surjective over the field~$A^\sim$. This is
      a contradiction to basic linear algebra. Hence $\Spec(A) \models \bot$,
      thus $1 = 0$ in $A$.
    \end{column}
    \begin{column}[t]{0.5\textwidth}
      \centering

@@

@@

  % XXX: \includegraphics[height=3em]{generic-freeness}
  \vspace*{-1em}

  \begin{varblock}{\textwidth}{Generic freeness$\phantom{p}$}
    \justifying
    Let~$B$ be an~$A$-algebra of finite type ($\cong A[X_1,\ldots,X_n]/\aaa$). \\
    Let~$M$ be a finitely generated~$B$-module ($\cong B^m/U$).

    If~$f = 0$ is the only element of~$A$ such that
    \begin{enumerate}
      \item $B[f^{-1}]$ and $M[f^{-1}]$ are free modules over $A[f^{-1}]$, \\[-1.6em]
      \item $A[f^{-1}] \to B[f^{-1}]$ is of finite presentation and \\[-1.6em]
      \item $M[f^{-1}]$ is finitely presented as a module over~$B[f^{-1}]$,
    \end{enumerate}
    then $1 = 0$ in~$A$.
  \end{varblock}
  \vspace*{-0.5em}

  \pause\raggedright\bigskip
  \textbf{Proof.} In the little Zariski topos it's \hil{not not} the case that % \\[-1.4em]
  \begin{enumerate}
    \item $B^\sim$ and $M^\sim$ are free modules over $A^\sim$,
    \item $A^\sim \to B^\sim$ is of finite presentation and
    \item $M^\sim$ is finitely presented as a module over $B^\sim$
  \end{enumerate}
  by basic linear algebra over the field $A^\sim$. The claim is precisely the
  external translation of this fact.
\end{frame}

@@

@@

  \begin{columns}[c]
    \begin{column}{0.43\textwidth}
      \begin{exampleblock}{}
        \justifying
        Let~$0 \to M' \to M \to M'' \to 0$ be a short exact sequence of
        modules. If~$M'$ and~$M''$ are finitely generated, so is~$M$.
      \end{exampleblock}
    \end{column}

    \scalebox{3}{$\Rightarrow$}
    \hspace*{0.5em}

    \begin{column}{0.47\textwidth}
      \begin{exampleblock}{}
        \justifying
        Let $0 \to \F' \to \F \to \F'' \to 0$ be a short exact sequence
        of sheaves of~$\O_X$-modules. If~$\F'$ and~$\F''$ are of finite type,
        so is~$\F$.
      \end{exampleblock}
    \end{column}
  \end{columns}
\end{frame}

@@

  \hil{Synthetic approach:} model schemes \hil{directly as sets} in a certain
  nonclassical set theory, the internal universe of the \mbox{\hil{big Zariski
  topos}} of a base scheme.
  \small
  \[ \{ (x,y,z) \? (\affl)^3 \,|\, x^n+y^n-z^n=0 \} \]
\end{frame}

\begin{frame}{The big Zariski topos}
  The \hil{big Zariski topos} $\Zar(S)$ of a scheme~$S$ is equivalently
  \begin{enumerate}
    \item the topos of sheaves over~$(\Aff/S)_{\mathrm{lofp}}$,
    \item the classifying topos of local rings over~$S$ or
    \item the classifying~$\Sh(S)$-topos of local~$\O_S$-algebras which are local
    over~$\O_S$.
  \end{enumerate}

  \begin{itemize}
    \item For an~$S$-scheme~$X$, its functor of points $\ull{X} =
    \Hom_S(\cdot,X)$ is an object of~$\Zar(S)$. It feels like \hil{the set of
    points} of~$X$.
    \item The ring $\affl$, given by $\affl(T) = \O_T(T)$,
    is a \hil{field} (nonzero implies invertible).
    \item \ \\[-1.2em]\mbox{$\ull{\mathbb{P}}^n = \llbracket \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\,
      x_0 \neq 0 \vee \cdots \vee x_n \neq 0 \}/(\affl)^\times \rrbracket$}
    \item $\ull{\Spec_S(\B)} = \llbracket \Hom_\affl(\B^\Zar,\affl) \rrbracket$
    \item $\ull{U} \hookrightarrow \ull{X}$ qc-open immersion iff
    \[ \Zar(S) \models \forall x\?\ull{X}\_ \bigvee_{n\geq0} \exists f_1,\ldots,f_n\?\affl\_ x \in \ull{U} \Leftrightarrow f_1 \neq 0 \vee \cdots \vee f_n \neq 0. \]
  \end{itemize}
\end{frame}

>>>>

\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\renewcommand{\AA}{\mathbb{A}}
\renewcommand{\C}{\mathcal{C}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\renewcommand{\G}{\mathcal{G}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\GG}{\mathbb{G}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\ZZ}{\mathbb{Z}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\fff}{\mathfrak{f}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\Zar}{\mathrm{Zar}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\newcommand{\Sch}{\mathrm{Sch}}
\newcommand{\Aff}{\mathrm{Aff}}
\newcommand{\LRS}{\mathrm{LRS}}
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\Spec}{\mathrm{Spec}}
\newcommand{\lra}{\longrightarrow}
\newcommand{\RelSpec}{\operatorname{Spec}}
\renewcommand{\_}{\mathpunct{.}}
\newcommand{\?}{\,{:}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\ull}[1]{\underline{#1}}
\newcommand{\affl}{\ensuremath{{\ull{\AA}^1}}}
\newcommand{\Ll}{\text{iff}}
\newcommand{\inv}{inv.\@}
\newcommand{\seq}{\vdash_{\!\!\!\vec x}}

@@

\newenvironment{indentblock}{%
  \list{}{\leftmargin\leftmargin}%
  \item\relax
}{%
  \endlist
}

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses}\end{minipage}}
\begin{frame}[c]
  \centering
  \medskip
  \hspace*{-3.9em}%
%  \includegraphics[height=0.3\textheight]{vichy-2}%
%  \includegraphics[height=0.3\textheight]{vichy-1}%
%  \includegraphics[height=0.3\textheight]{vichy-4}%
%  \includegraphics[height=0.3\textheight]{vichy-5}%

  \hil{How topos theory can help algebra} \\

@@

  \bigskip
  \bigskip

@@

      \hil{Background on the internal language}
      \medskip

    \end{column}
    \hspace*{0.5em}
    \begin{column}{0.35\textwidth}
      \centering
      \bignumber{2}

      \hil{Applications in commutative algebra}
      \medskip

@@

      \hil{The mystery of nongeometric sequents}
      \medskip

@@

  \scriptsize
  Ingo Blechschmidt (MPI Leipzig) \\
  Toposes in Como \\
  June 29th, 2018
  \par
\end{frame}}

\section{Motivation}

@@

  \includegraphics[height=3em]{generic-freeness}
  \vspace*{-1em}

@@

  \pause\center
  \begin{minipage}{0.9\textwidth}
    \textbf{Proof.} See [Stacks Project, Tag 051Q].
  \end{minipage}
\end{frame}

\section{Internal language}

@@

  \vspace*{-1em}
  \begin{columns}
    \def\insertblocktitle{}
    \begin{column}{0.25\textwidth}\usebeamertemplate{block begin}
      \centering
      $\Set \models \varphi$ \\
      ``$\varphi$ holds in the \\ usual sense.''
      \usebeamertemplate{block end}\end{column}

    \begin{column}{0.25\textwidth}\usebeamertemplate{block begin}
      \centering
      $\Sh(X) \models \varphi$ \\
      ``$\varphi$ holds continuously.''
      \usebeamertemplate{block end}\end{column}

    \begin{column}{0.25\textwidth}\usebeamertemplate{block begin}
      \centering
      $\Eff \models \varphi$ \\
      ``$\varphi$ holds \\ computably.''
      \usebeamertemplate{block end}\end{column}
  \end{columns}
  \medskip

@@

\begin{frame}{Internalizing parameter-dependence}
  \justifying
  Let~$X$ be a space. A continuous family~$(f_x)_{x \in X}$ of continuous
  functions (that is, a continuous function~$f : X \times \RR \to \RR$;
  $f_x(a) = f(x,a)$) induces an endomorphism of the sheaf~$\C$ of continuous
  functions:
  \[
    \bar f : \C \longrightarrow \C,\
    \text{on $U$:}\ s \longmapsto (x \mapsto f_x(s(x))).
  \]

  \vspace*{-1em}

  \begin{itemize}
    \item $\Sh(X) \models \speak{The set $\C$ is the set of (Dedekind) reals}$.
    \item $\Sh(X) \models \speak{The function $\bar f : \RR \to \RR$ is continuous}$. \bigskip
    \item Iff $f_x(-1) < 0$ for all $x$, then $\Sh(X) \models \bar f(-1) < 0$.
    \item Iff $f_x(+1) > 0$ for all $x$, then $\Sh(X) \models \bar f(+1) > 0$.
    \item Iff all $f_x$ are increasing, then $\Sh(X) \models \speak{$\bar f$ is
    increasing}$. \bigskip
    \item \ \\[-1.2em]\mbox{Iff there is an open cover~$X = \bigcup_i U_i$ such
    that for each~$i$ there is a} \mbox{continuous function~$s : U_i \to \RR$
    with~$f_x(s(x)) = 0$ for all~$x \in U_i$,} then $\Sh(X) \models \exists s
    \? \RR\_ \bar f(s) = 0$.
  \end{itemize}
\end{frame}

\section{The little Zariski topos}

\begin{frame}{The little Zariski topos}
  Let~$A$ be a ring. Its \hil{little Zariski topos} is equivalently
  \begin{enumerate}
    \item the classifying topos of \hil{local localizations} of $A$,
    \item the classifying locale of \hil{prime filters} of $A$,
    \item the locale given by the frame of \hil{radical ideals} of $A$,
    \item the topos of sheaves over the poset $A$ with $f \preceq g$ iff $f \in \sqrt{(g)}$ and with $(f_i \to f)_i$ deemed covering iff $f \in \sqrt{(f_i)_i}$ or
    \item the topos of sheaves over $\Spec(A)$.
  \end{enumerate}
  Its associated topological space of points is the \hil{classical spectrum}
  \[ \{ \fff \subseteq A \,|\, \text{$\fff$ prime filter} \} + \text{Zariski topology}. \]
  It has \hil{enough points} if the Boolean Prime Ideal Theorem holds.

@@

  \pause
  \only<2>{\begin{center}
    \includegraphics[width=0.9\textwidth]{hartshorne-tilde-construction} \\
    \footnotesize
    Robin Hartshorne. Algebraic Geometry. 1977.
  \end{center}}
  \pause

  \vspace*{-1.5em}

  \begin{columns}
    \begin{column}{0.48\textwidth}
      \begin{varblock}{\textwidth}{}
        \justifying
        Assuming the Boolean prime ideal theorem, a geometric sequent
        ``$\forall \ldots \forall\_ (\cdots \Longrightarrow \cdots\!\,)$'',
        where the two subformulas may not contain~``$\Rightarrow$'' and~``$\forall$'',
        holds for~$M^\sim$ iff it holds for all stalks~$M_\ppp$.
      \end{varblock}

      \vspace*{-1.5em}

      \begin{varblock}{\textwidth}{}
        $M^\sim$ inherits any property of~$M$ which is \hil{localization-stable}.
      \end{varblock}
    \end{column}

    \begin{column}{0.5\textwidth}
      \vspace*{1.5em}

      If $A$ is reduced ($x^n = 0 \Rightarrow x = 0$):

      \vspace*{-1.0em}

      \setbeamercolor{block body}{bg=red!30}
      \setbeamercolor{structure}{fg=purple}
      \begin{varblock}{\textwidth}{}
        $A^\sim$ is a \hil{field}
        (nonunits are zero).

@@

  \visible<4-5>{\begin{tikzpicture}[overlay]
    \draw[fill=white, draw=white, opacity=0.95] (-1,0) rectangle (\paperwidth,7.4);
    \node[anchor=south west,inner sep=0] (image) at (0,0.8) {\vbox{
      \only<4>{
	\centering
	\includegraphics[width=0.9\textwidth]{tierney-on-the-spectrum-of-a-ringed-topos} \\
	\footnotesize
	Miles Tierney. On the spectrum of a ringed topos. 1976.
      }

      \only<5>{
	\begin{varblock}{\textwidth}{Complexity reduction}
	  The external meaning of
	  \[
	    \Spec(A) \models
	      \speak{$A^\sim[X_1,\ldots,X_n]$ is anonymously Noetherian}
	  \]
	  is:
	  \medskip

	  \begin{indentblock}
	  For any element~$f \in A$ and any (not necessarily quasicoherent) sheaf of
	  ideals~$\J \hookrightarrow A^\sim[X_1,\ldots,X_n]|_{D(f)}$: If
	  \begin{indentblock}
	  for any element~$g \in A$ the condition that
	  \begin{indentblock}
	  the sheaf~$\J$ is of finite type on~$D(g)$
	  \end{indentblock}
	  implies that~$g = 0$,
	  \end{indentblock}
	  then~$f = 0$.
	  \end{indentblock}
	\end{varblock}

	\vspace*{-2em}
      }
    }};
  \end{tikzpicture}}
\end{frame}

@@

  \only<1>{\begin{columns}[t]
    \begin{column}[t]{0.47\textwidth}
      \centering

@@

      \justifying
      \textbf{Proof.} The matrix is surjective over the field~$A^\sim$.
      This is a contradiction to basic linear algebra. Hence $\Spec(A) \models
      \bot$, thus $1 = 0$ in $A$.
    \end{column}
    \begin{column}[t]{0.47\textwidth}
      \centering

@@

  \vspace*{-5.3em}\hspace*{21em}\begin{minipage}{3cm}
    $\xymatrixcolsep{3.5pc}\xymatrixrowsep{1.2pc}\xymatrix{
      & \ar@{-}[d]^{\substack{\text{finitely}\\\text{generated}}} M \\
      A \ar[r]_{\text{of finite type}} & B
    }$
  \end{minipage}
  \vspace*{-1em}

@@

  \scalebox{0.9}{\begin{minipage}{1\textwidth}
    \textbf{Proof.} In the little Zariski topos it's \hil{not not} the case that % \\[-1.4em]
    \begin{enumerate}
      \item $B^\sim$ and $M^\sim$ are free modules over $A^\sim$,
      \item \vspace*{-0.3em}$A^\sim \to B^\sim$ is of finite presentation and
      \item \vspace*{-0.3em}$M^\sim$ is finitely presented as a module over $B^\sim$,
    \end{enumerate}
    by basic linear algebra over the field $A^\sim$. The claim is precisely the
    external translation of this fact.
  \end{minipage}}
\end{frame}

@@

\section{Algebraic geometry}

@@

  \begin{columns}[c]
    \begin{column}{0.47\textwidth}
      \begin{exampleblock}{}
        \justifying
        Let $0 \to \F' \to \F \to \F'' \to 0$ be a short exact sequence
        of sheaves of~$\O_X$-modules. If~$\F'$ and~$\F''$ are of finite type,
        so is~$\F$.
      \end{exampleblock}
    \end{column}

    \begin{column}{0.1\textwidth}
      \vspace*{0.7em}
      \scalebox{3}{$\Leftarrow$}
    \end{column}

    \begin{column}{0.44\textwidth}
      \begin{exampleblock}{}
        \justifying
        Let~$0 \to M' \to M \to M'' \to 0$ be a short exact sequence of
        modules. If~$M'$ and~$M''$ are finitely generated, so is~$M$.
      \end{exampleblock}
    \end{column}
  \end{columns}
\end{frame}

@@

  \hil{Synthetic approach:} model schemes \hil{directly as sets} in
  the internal universe of the \mbox{\hil{big Zariski topos}} of a base scheme.
  \small
  \[ \{ (x,y,z) \? (\affl)^3 \,|\, x^n+y^n-z^n=0 \} \]
\end{frame}

\begin{frame}{The big Zariski topos}
  Let~$S$ be a fixed base scheme.
  \begin{varblock}{0.9\textwidth}{Definition}
    The \hil{big Zariski topos} $\Zar(S)$ of a scheme~$S$ is equivalently
    \begin{enumerate}
      \item the topos of sheaves over~$(\Aff/S)_{\mathrm{lofp}}$,
      \item the classifying topos of local rings over~$S$ or
      \item the classifying~$\Sh(S)$-topos of local~$\O_S$-algebras which are
      local over~$\O_S$.
    \end{enumerate}
  \end{varblock}

  \begin{itemize}
    \item For an~$S$-scheme~$X$, its functor of points $\ull{X} =
    \Hom_S(\cdot,X)$ is an object of~$\Zar(S)$. It feels like \hil{the set of
    points} of~$X$.
    \item In particular, there is the ring object~$\affl$ with~$\affl(T) =
    \O_T(T)$.
    \item This ring object is a \hil{field}: nonzero implies
    invertible. \\{} [Kock 1976]
  \end{itemize}
\end{frame}

\begin{frame}{Synthetic constructions}
  \small
  $\hil{$\boldsymbol{\mathbb{A}^n}$} = (\affl)^n = \affl \times \cdots \times \affl$

  $\begin{array}{@{}r@{}c@{}l@{}}
    \hil{$\boldsymbol{\mathbb{P}^n}$} &\phantom{}=\phantom{}& \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\, x_0 \neq 0 \vee
    \cdots \vee x_n \neq 0 \}/(\affl)^\times \\
    &\cong& \text{set of one-dimensional subspaces of~$(\affl)^{n+1}$} \\
    &&\qquad \text{(with~$\O(-1) = (\ell)_{\ell \? \mathbb{P}^n}$, $\O(1) =
    (\ell^\vee)_{\ell \? \mathbb{P}^n}$)}
  \end{array}$

  $\hil{$\boldsymbol{\Spec(R)}$} = \Hom_{\mathrm{Alg}(\affl)}(R, \affl) =
    \text{set of $\affl$-valued points of $R$}$

  $\hil{$\boldsymbol{TX}$} = X^\Delta$, where $\Delta = \{ \varepsilon \? \affl \,|\, \varepsilon^2 = 0 \}$

  A subset $U \subseteq X$ is \hil{qc-open} if and only if for any $x : X$
  there exist $f_1,\ldots,f_n \? \affl$ such that $x \in U \Longleftrightarrow
  \exists i\_ f_i \neq 0$.

  A \hil{synthetic affine scheme} is a set which is in bijection
  with~$\Spec(R)$ for some synthetically quasicoherent~$\affl$-algebra~$R$.

  A \hil{synthetic scheme} is a set which can be covered by finitely many
  qc-open synthetic affine schemes~$U_i$ such that the intersections~$U_i \cap
  U_j$ can be covered by finitely many qc-open synthetic affine schemes.
\end{frame}

\section{The mystery of nongeometric sequents}

\begin{frame}{Properties of the affine line}
  \begin{itemize}
    \item $\affl$ is a local ring:
    \[ 1 \neq 0 \qquad\qquad \text{$x + y$ \inv} \Longrightarrow \text{$x$ \inv} \vee
    \text{$y$ \inv} \]
    \item $\affl$ is a field:
    \begin{align*}
      \neg(x = 0) &\Longleftrightarrow \text{$x$ invertible \quad [Kock 1976]} \\
      \neg(\text{$x$ invertible}) &\Longleftrightarrow \text{$x$ nilpotent}
    \end{align*}
    \item $\affl$ satisfies the axiom of microaffinity: Any map $f : \Delta \to
    \affl$ is of the form~$f(\varepsilon) = a + b \varepsilon$ for unique
    values~$a,b \? \affl$, where~$\Delta = \{ \varepsilon \? \affl \,|\,
    \varepsilon^2 = 0 \}$.
    \item Any map $\affl \to \affl$ is a polynomial.
    \item $\affl$ is anonymously algebraically closed: Any monic polynomial
    does \emph{not not} have a zero.
  \end{itemize}
\end{frame}

\begin{frame}{Synthetic quasicoherence}
  Recall~$\Spec(R) = \Hom_{\mathrm{Alg}(\affl)}(R, \affl)$ and consider the statement
  \[ \text{``the canonical map~$
    \begin{array}[t]{rcl}
      R &\longrightarrow& (\affl)^{\Spec(R)} \\
      f &\longmapsto& (\alpha \mapsto \alpha(f))
    \end{array}
  $ is bijective''}.
  \]
  \vspace*{-1em}

  \begin{itemize}
    \item True for~$R = \affl[X]/(X^2)$ (microaffinity).
    \item True for~$R = \affl[X]$ (every function is a polynomial).
    \item True for \hil{any} finitely presented~$\affl$-algebra~$R$.
  \end{itemize}

  Any known property of~$\affl$ follows from this \\
  \hil{synthetic quasicoherence}.

  \centering
  \hil{The mystery of nongeometric sequents}

% \emph{Example.} Let~$x \? \affl$ such that~$x \neq 0$. Set~$R = \affl/(x)$.
% Then~$\Spec(R) = \emptyset$. Thus~$(\affl)^{\Spec(R)}$ is a singleton.
% Hence~$R = 0$. Therefore~$x$ is invertible.
\end{frame}

\begin{frame}{Classifying toposes in algebraic geometry}
  \small
  \centering
  \begin{tabular}{ll}
    \toprule
    (Big) topos & classified theory \\\midrule
    Zariski & local rings [Hakim 1972] \\[0.6em]
    étale & separably closed local rings [Hakim 1972, Wraith 1979] \\[0.6em]
    \usebeamercolor[fg]{item}{fppf} & \usebeamercolor[fg]{item}{fppf-local rings} \\
    & \textcolor{red!90}{(conjecturally: algebraically closed local rings)} \\[0.6em]
    \textcolor{red!80}{ph} & \textcolor{red!80}{?? (conjecturally: algebraically closed valuation} \\
    & \qquad \textcolor{red!80}{rings validating the projective Nullstellensatz)} \\[0.6em]
    \usebeamercolor[fg]{item}{surjective} & \usebeamercolor[fg]{item}{algebraically closed geometric fields} \\[0.6em]
    \textcolor{red!80}{$\neg\neg$} & \textcolor{red!80}{?? (conjecturally: algebraically closed geometric} \\
    & \qquad \textcolor{red!80}{fields are integral over the base)} \\[0.6em]
    \usebeamercolor[fg]{item}{infinitesimal} & \usebeamercolor[fg]{item}{XXX [Hutzler 2018]} \\[0.6em]
    \textcolor{red!100}{crystalline} & \textcolor{red!100}{??} \\
    \bottomrule
  \end{tabular}
\end{frame}
